---
title: "Week 2 - Homework"
author: "STAT 420, Summer 2021, D. Unger"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---

***

## Exercise 1 (Using `lm`)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Suppose we would like to understand the size of a cat's heart based on the body weight of a cat. Fit a simple linear model in `R` that accomplishes this task. Store the results in a variable called `cat_model`. Output the result of calling `summary()` on `cat_model`.

```{r 1a}
library(MASS)
cats <- MASS::cats
Hwt <- cats$Hwt
Bwt <- cats$Bwt
cat_model <- lm(Hwt ~ Bwt, data = cats)
summary(cat_model)
```

**(b)** Output only the estimated regression coefficients. Interpret $\hat{\beta_0}$ and $\beta_1$ in the *context of the problem*. Be aware that only one of those is an estimate.

```{r 1b}
cat_model$coefficients
```
$\hat{\beta_0}$ is the estimated mean of cat heart weight when cat weight is zero
$\beta_1$ is the mean of increase of cat heart weight when cat weight increase by 1

**(c)** Use your model to predict the heart weight of a cat that weights **3.1** kg. Do you feel confident in this prediction? Briefly explain.

```{r 1c}
predict(cat_model, newdata = data.frame(Bwt = 3.1))
```
Yes because 3.1 is in the range if known sample data.

**(d)** Use your model to predict the heart weight of a cat that weights **1.5** kg. Do you feel confident in this prediction? Briefly explain.

```{r }
predict(cat_model, newdata = data.frame(Bwt = 1.5))
```
No. 1.5 is outside of the sample range and extrapolating.

**(e)** Create a scatterplot of the data and add the fitted regression line. Make sure your plot is well labeled and is somewhat visually appealing.

```{r 1e}
plot(Hwt ~ Bwt, data = cats,
     main = "Heart weight vs Body weight of cats",
     xlab = "Body Weight",
     ylab = "Heart Weight",
     col = "grey")

abline(cat_model, lwd = 3, col = "darkorange")
```

**(f)** Report the value of $R^2$ for the model. Do so directly. Do not simply copy and paste the value from the full output in the console after running `summary()` in part **(a)**.

```{r 1f}
summary(cat_model)['r.squared']
```


***

## Exercise 2 (Writing Functions)

This exercise is a continuation of Exercise 1.

**(a)** Write a function called `get_sd_est` that calculates an estimate of $\sigma$ in one of two ways depending on input to the function. The function should take three arguments as input:

- `fitted_vals` - A vector of fitted values from a model
- `actual_vals` - A vector of the true values of the response
- `mle` - A logical (`TRUE` / `FALSE`) variable which defaults to `FALSE`

The function should return a single value:

- $s_e$ if `mle` is set to `FALSE`.
- $\hat{\sigma}$ if `mle` is set to `TRUE`.

```{r 2a}
get_sd_est = function(fitted_vals, actual_vals, mle = FALSE) {
  e <- actual_vals - fitted_vals
  n <- length(e)
  if(mle == FALSE) {
    
    s2_e <- sum(e ^ 2) / (n - 2)
    s_e <- sqrt(s2_e)
    return(s_e)
  } else {
    sigma_hat <- sqrt(sum(e ^ 2) / n)
    return(sigma_hat)
  }
}
```

**(b)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `FALSE`. Explain the resulting estimate in the context of the model.

```{r 2b}
  get_sd_est(residuals(cat_model), rep(0, times= length(residuals(cat_model))))
```
This is the variance of residuals (cat heart weight) using least squares method.
**(c)** Run the function `get_sd_est` on the residuals from the model in Exercise 1, with `mle` set to `TRUE`. Explain the resulting estimate in the context of the model. Note that we are trying to estimate the same parameter as in part **(b)**.

```{r 2c}
get_sd_est(residuals(cat_model), rep(0, times= length(residuals(cat_model))), mle = TRUE)
```
This is the variance of residuals (cat heart weight) using maximum likelyhood estimation method.


**(d)** To check your work, output `summary(cat_model)$sigma`. It should match at least one of **(b)** or **(c)**.

```{r 2d}
summary(cat_model)$sigma
```

***

## Exercise 3 (Simulating SLR)

Consider the model

\[
Y_i = 5 + -3 x_i + \epsilon_i
\]

with 

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 10.24)
\]

where $\beta_0 = 5$ and $\beta_1 = -3$.

This exercise relies heavily on generating random observations. To make this reproducible we will set a seed for the randomization. Alter the following code to make `birthday` store your birthday in the format: `yyyymmdd`. For example, [William Gosset](https://en.wikipedia.org/wiki/William_Sealy_Gosset), better known as *Student*, was born on June 13, 1876, so he would use:

```{r}
birthday = 19880210
set.seed(birthday)
```

**(a)** Use `R` to simulate `n = 25` observations from the above model. For the remainder of this exercise, use the following "known" values of $x$.

```{r 3a}
sim_slr = function(x, beta_0 = 10, beta_1 = 5, sigma = 1) {
  n = length(x)
  epsilon = rnorm(n, mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  data.frame(predictor = x, response = y)
}

x = runif(n = 25, 0, 10)
sim_data = sim_slr(x = x, beta_0 = 5, beta_1 = -3, sigma = sqrt(10.24))


```

You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Store the data frame this function returns in a variable of your choice. Note that this function calls $y$ `response` and $x$ `predictor`.

**(b)** Fit a model to your simulated data. Report the estimated coefficients. Are they close to what you would expect? Briefly explain.


```{r 3b}
sim_data_response = sim_data$response
lm_sim = lm(sim_data_response ~ x)
coef(lm_sim)

```
No they are not close to my expectation since the number of samples are small.


**(c)** Plot the data you simulated in part **(a)**. Add the regression line from part **(b)** as well as the line for the true model. Hint: Keep all plotting commands in the same chunk.


```{r 3c}
plot(sim_data_response ~ x,
     main = "Simulationof SLR",
     xlab = "x",
     ylab = "Simulated values",
     col = "grey"
     )
abline(lm_sim, lwd = 3, col = "darkorange")

```
**(d)** Use `R` to repeat the process of simulating `n = 25` observations from the above model $1500$ times. Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. Some hints:

- Consider a `for` loop.
- Create `beta_hat_1` before writing the `for` loop. Make it a vector of length $1500$ where each element is `0`.
- Inside the body of the `for` loop, simulate new $y$ data each time. Use a variable to temporarily store this data together with the known $x$ data as a data frame.
- After simulating the data, use `lm()` to fit a regression. Use a variable to temporarily store this output.
- Use the `coef()` function and `[]` to extract the correct estimated coefficient.
- Use `beta_hat_1[i]` to store in elements of `beta_hat_1`.
- See the notes on [Distribution of a Sample Mean](http://daviddalpiaz.github.io/appliedstats/introduction-to-r.html#distribution-of-a-sample-mean) for some inspiration.

You can do this differently if you like. Use of these hints is not required.


```{r 3d}
beta_hat_1 <- rep(0, times = 1500)
for(i in 1: 1500) {
  sim_data = sim_slr(x = x, beta_0 = 5, beta_1 = -3, sigma = sqrt(10.24))
  sim_data_response = sim_data$response
  lm_sim = lm(sim_data_response ~ x)
  b1s = coef(lm_sim)['x']
  beta_hat_1[i] = b1s
}

```

**(e)** Report the mean and standard deviation of `beta_hat_1`. Do either of these look familiar?


```{r 3e}
mean(beta_hat_1)
sd(beta_hat_1)

```
The mean if beta_hat_1 roughly equals to $\beta_1 = -3$
**(f)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.


```{r }
hist(beta_hat_1)
```
The shape is a normal distribution with mean around $\beta_1 = -3$
***

## Exercise 4 (Be a Skeptic)

Consider the model

\[
Y_i = 3 + 0 \cdot x_i + \epsilon_i
\]

with

\[
\epsilon_i \sim N(\mu = 0, \sigma^2 = 4)
\]

where $\beta_0 = 3$ and $\beta_1 = 0$.

Before answering the following parts, set a seed value equal to **your** birthday, as was done in the previous exercise.

```{r}
birthday = 19880210
set.seed(birthday)
```

**(a)** Use `R` to repeat the process of simulating `n = 75` observations from the above model $2500$ times. For the remainder of this exercise, use the following "known" values of $x$.

```{r}
x = runif(n = 75, 0, 10)
beta_hat_1 <- rep(0, times = 2500)
for(i in 1: 2500) {
  sim_data = sim_slr(x = x, beta_0 = 3, beta_1 = 0, sigma = 2)
  sim_data_response = sim_data$response
  lm_sim = lm(sim_data_response ~ x)
  b1s = coef(lm_sim)['x']
  beta_hat_1[i] = b1s
}
```

Each time fit a SLR model to the data and store the value of $\hat{\beta_1}$ in a variable called `beta_hat_1`. You may use [the `sim_slr ` function provided in the text](http://daviddalpiaz.github.io/appliedstats/simple-linear-regression.html#simulating-slr). Hint: Yes $\beta_1 = 0$.

**(b)** Plot a histogram of `beta_hat_1`. Comment on the shape of this histogram.

```{r 4b}
hist(beta_hat_1)
```
The shape is a normal distribution with mean equas to 0.

**(c)** Import the data in [`skeptic.csv`](skeptic.csv) and fit a SLR model. The variable names in `skeptic.csv` follow the same convention as those returned by `sim_slr()`. Extract the fitted coefficient for $\beta_1$.


```{r 4c}
skcsv <- read_csv('skeptic.csv')
sk_lm <- lm(response ~ predictor, data = skcsv)
sk_beta_hat_1 <- coef(sk_lm)['predictor']
sk_beta_hat_1

```


**(d)** Re-plot the histogram from **(b)**. Now add a vertical red line at the value of $\hat{\beta_1}$ in part **(c)**. To do so, you'll need to use `abline(v = c, col = "red")` where `c` is your value.

```{r 4d}
hist(beta_hat_1)
abline(v = sk_beta_hat_1, col = "red")

```


**(e)** Your value of $\hat{\beta_1}$ in **(c)** should be negative. What proportion of the `beta_hat_1` values is smaller than your $\hat{\beta_1}$? Return this proportion, as well as this proportion multiplied by `2`.


```{r 4e}
p_small <- length(beta_hat_1[beta_hat_1 < sk_beta_hat_1]) / length(beta_hat_1)
p_small
p_small * 2
```


**(f)** Based on your histogram and part **(e)**, do you think the [`skeptic.csv`](skeptic.csv) data could have been generated by the model given above? Briefly explain.

No. Only a very small percentage of sample data is smaller than the mean of skeptic model. It is unlikely that skeptic data is generated by the model given above.

***

## Exercise 5 (Comparing Models)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will perform some data cleaning before proceeding.

```{r}
library(mlbench)
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

We have:

- Loaded the data from the package
- Subset the data to relevant variables
    - This is not really necessary (or perhaps a good idea) but it makes the next step easier
- Given variables useful names
- Removed any observation with missing values
    - This should be given much more thought in practice

For this exercise we will define the "Root Mean Square Error" of a model as

\[
\text{RMSE} = \sqrt{\frac{1}{n} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2}.
\]

**(a)** Fit three SLR models, each with "ozone" as the response. For the predictor, use "wind speed," "humidity percentage," and "temperature" respectively. For each, calculate $\text{RMSE}$ and $R^2$. Arrange the results in a markdown table, with a row for each model. Suggestion: Create a data frame that stores the results, then investigate the `kable()` function from the `knitr` package.


```{r 5a}
library(knitr)
wind_lm <- lm(ozone ~ wind, data = Ozone)
hum_lm <- lm(ozone ~ humidity, data = Ozone)
temp_lm <- lm(ozone ~ temp, data = Ozone)

get_rmse = function(e, mle = FALSE) {
  n <- length(e)
 
    rmse <- sqrt(sum(e ^ 2) / n)
    return(rmse)
  
}

wind_lm_rmse <- get_rmse(resid(wind_lm))
hum_lm_rmse <- get_rmse(resid(hum_lm))
temp_lm_rmse <- get_rmse(resid(temp_lm))

wind_lm_r2 <- summary(wind_lm)$r.squared
hum_lm_r2 <- summary(hum_lm)$r.squared
temp_lm_r2 <- summary(temp_lm)$r.squared

rmse <- c(wind_lm_rmse, hum_lm_rmse, temp_lm_rmse)
r2 <- c(wind_lm_r2, hum_lm_r2, temp_lm_r2)
rmse_r2_df = data.frame(rmse, r2)
row.names(rmse_r2_df) <- c('wind', 'humidity', 'temp')

kable(rmse_r2_df,  caption = "Ozone vs Wind/Humidity/Temp")

```


**(b)** Based on the results, which of the three predictors used is most helpful for predicting ozone readings? Briefly explain.
Temperature is most useful. It has highest R2 and lowest RMSE. Shows the data fits better in Ozone vs Temperature.


***



