---
title: "Week 3 - Homework"
author: "STAT 420, Summer 2021, D. Unger"
date: ''
output:
  html_document: 
    theme: readable
    toc: yes  
  pdf_document: default
urlcolor: cyan
---


***

## Exercise 1 (Using `lm` for Inference)

For this exercise we will use the `cats` dataset from the `MASS` package. You should use `?cats` to learn about the background of this dataset.

**(a)** Fit the following simple linear regression model in `R`. Use heart weight as the response and body weight as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `cat_model`. Use a $t$ test to test the significance of the regression. Report the following:
```{r 1a}
library(MASS)
cats <- MASS::cats
cat_model <- lm(Hwt ~ Bwt, data = cats)
summary(cat_model)
```

- The null and alternative hypotheses
Null hypothesis - \[H_0: \beta_1 = 0\]
Alternative hypothesis - \[H_1: \beta_1 \neq 0\]

- The value of the test statistic
```{r }
beta_0_hat_t <- summary(cat_model)$coefficients["(Intercept)","t value"]
beta_1_hat_t <- summary(cat_model)$coefficients["Bwt","t value"]
beta_0_hat_t
beta_1_hat_t
```


- The p-value of the test

For \[\beta_1\]
p value = <2e-16

- A statistical decision at $\alpha = 0.05$
Reject the null hypothesis.

- A conclusion in the context of the problem
Heart weight has a significant linear relationship with body weight.


When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**(b)** Calculate a 95% confidence interval for $\beta_1$. Give an interpretation of the interval in the context of the problem.
```{r 2b}
confint(cat_model, level = 0.95)['Bwt',]
```
We are 95% confident that the true change in mean heart weight for an increase in body weight of 1 kg is between 3.539343 and 4.528782.


**(c)** Calculate a 90% confidence interval for $\beta_0$. Give an interpretation of the interval in the context of the problem.
```{r 1c}
confint(cat_model, level = 0.95)['(Intercept)',]
```
We are 90% confident that the true change in heart weight when the body weight is 0 is between -1.725163  1.011838.

**(d)** Use a 90% confidence interval to estimate the mean heart weight for body weights of 2.1 and 2.8 kilograms. Which of the two intervals is wider? Why?
```{r 1d}
body_weights <- c(2.1, 2.8)
bw_df <- data.frame(Bwt = body_weights)
predict(cat_model,  newdata = bw_df, 
        interval = c("confidence"), level = 0.90)
```

2.1 kilograms is wider. It's father from \[\bar{x}\].

**(e)** Use a 90% prediction interval to predict the heart weight for body weights of 2.8 and 4.2 kilograms.
```{r 1e}
body_weights <- c(2.8, 4.2)
bw_df <- data.frame(Bwt = body_weights)
predict(cat_model,  newdata = bw_df, 
        interval = c("prediction"), level = 0.90)
```

**(f)** Create a scatterplot of the data. Add the regression line, 95% confidence bands, and 95% prediction bands.
```{r 1f}
hwt_grid = seq(min(cats$Bwt), max(cats$Bwt), by = 0.01)
dist_ci_band = predict(cat_model, 
                       newdata = data.frame(Bwt = hwt_grid), 
                       interval = "confidence", level = 0.99)
dist_pi_band = predict(cat_model, 
                       newdata = data.frame(Bwt = hwt_grid), 
                       interval = "prediction", level = 0.99) 

plot(Hwt ~ Bwt, data = cats,
     xlab = "Body Weight (In kilograms)",
     ylab = "Heart Weight(In Kilograms)",
     main = "Heart Weight vs Body Weight",
     pch  = 20,
     cex  = 2,
     col  = "grey",
     ylim = c(min(dist_pi_band), max(dist_pi_band)))
abline(cat_model, lwd = 5, col = "darkorange")

lines(hwt_grid, dist_ci_band[,"lwr"], col = "dodgerblue", lwd = 3, lty = 2)
lines(hwt_grid, dist_ci_band[,"upr"], col = "dodgerblue", lwd = 3, lty = 2)
lines(hwt_grid, dist_pi_band[,"lwr"], col = "dodgerblue", lwd = 3, lty = 3)
lines(hwt_grid, dist_pi_band[,"upr"], col = "dodgerblue", lwd = 3, lty = 3)
points(mean(cats$Hwt), mean(cats$Bwt), pch = "+", cex = 3)
```

**(g)** Use a $t$ test to test:

- $H_0: \beta_1 = 4$
- $H_1: \beta_1 \neq 4$

Report the following:

- The value of the test statistic
```{r 1.g.1}
beta_hat <- summary(cat_model)$coefficients[2,1]
beta_1_se <- summary(cat_model)$coefficients[2, 2]
beta_1_0 <- 4
beta_1_hat_t <- (beta_hat - beta_1_0) / beta_1_se
beta_1_hat_t
```
- The p-value of the test
```{r 1.g.2}
2 * pt(-abs(beta_1_hat_t), df = nrow(cats) - 2)
```
- A statistical decision at $\alpha = 0.05$

Fail to reject the \[H_0\].
When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***

## Exercise 2 (More `lm` for Inference)

For this exercise we will use the `Ozone` dataset from the `mlbench` package. You should use `?Ozone` to learn about the background of this dataset. You may need to install the `mlbench` package. If you do so, do not include code to install the package in your `R` Markdown document.

For simplicity, we will re-perform the data cleaning done in the previous homework.

```{r}
data(Ozone, package = "mlbench")
Ozone = Ozone[, c(4, 6, 7, 8)]
colnames(Ozone) = c("ozone", "wind", "humidity", "temp")
Ozone = Ozone[complete.cases(Ozone), ]
```

**(a)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and wind speed as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `ozone_wind_model`. Use a $t$ test to test the significance of the regression. Report the following:
```{r 2.a}
ozone_wind_model <- lm(ozone ~ wind, data = Ozone)
summary(ozone_wind_model)
```
- The null and alternative hypotheses
  Null - $H_0: \beta_1 = 0$
  Alternative - $H_1: \beta_1 \neq 0$
  
- The value of the test statistic
For \[\beta_0\] and \[\beta_1\]
```{r }
summary(ozone_wind_model)$coefficients[1,3]
summary(ozone_wind_model)$coefficients[2,3]
```
- The p-value of the test
```{r }
summary(ozone_wind_model)$coefficients[1,4]
summary(ozone_wind_model)$coefficients[2,4]
```

- A statistical decision at $\alpha = 0.01$
Failed to reject null hypotheis.

- A conclusion in the context of the problem
No significant linear relation between ozone and wind.

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

**(b)** Fit the following simple linear regression model in `R`. Use the ozone measurement as the response and temperature as the predictor. 

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Store the results in a variable called `ozone_temp_model`. Use a $t$ test to test the significance of the regression. Report the following:
```{r 2.b}
ozone_temp_model <- lm(ozone ~ temp, data = Ozone)
```
- The null and alternative hypotheses
  Null - $H_0: \beta_1 = 0$
  Alternative - $H_1: \beta_1 \neq 0$

- The value of the test statistic
```{r }
summary(ozone_temp_model)$coefficients[1,3]
summary(ozone_temp_model)$coefficients[2,3]
```

- The p-value of the test
```{r }
summary(ozone_temp_model)$coefficients[1,4]
summary(ozone_temp_model)$coefficients[2,4]
```

- A statistical decision at $\alpha = 0.01$
Reject the null hypothesis.

- A conclusion in the context of the problem
There's significant linear relationship between temperature and ozone, with significance level 0.01.

When reporting these, you should explicitly state them in your document, not assume that a reader will find and interpret them from a large block of `R` output.

***

## Exercise 3 (Simulating Sampling Distributions)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = -5$
- $\beta_1 = 3.25$
- $\sigma^2 = 16$

We will use samples of size $n = 50$.

**(a)** Simulate this model $2000$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_0$ and $\hat{\beta}_1$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 19880210
set.seed(birthday)
n = 50
x = seq(0, 10, length = n)
```

```{r}
sim_slr = function(x, beta_0 = 10, beta_1 = 5, sigma = 1) {
  n = length(x)
  epsilon = rnorm(n, mean = 0, sd = sigma)
  y = beta_0 + beta_1 * x + epsilon
  data.frame(predictor = x, response = y)
}
beta_0_hats <- rep(0, times=2000)
beta_1_hats <- rep(0, times= 2000)

for (i in 1:2000) {
  sim_res <- sim_slr(x = x, beta_0 = -5, beta_1 = 3.25, sigma = 4)
  linear_model <- lm(response ~ predictor, data = sim_res)
  
  beta_0_hats[i] <- summary(linear_model)$coefficients["(Intercept)", "Estimate"]
  beta_1_hats[i] <- summary(linear_model)$coefficients["predictor", "Estimate"]
}
```

**(b)** Create a table that summarizes the results of the simulations. The table should have two columns, one for $\hat{\beta}_0$ and one for $\hat{\beta}_1$. The table should have four rows:

- A row for the true expected value given the known values of $x$
- A row for the mean of the simulated values
- A row for the true standard deviation given the known values of $x$
- A row for the standard deviation of the simulated values

```{r}
sigma = 4
Sxx = sum((x - mean(x)) ^ 2)
beta_0_mean <- mean(beta_0_hats)
beta_1_mean <- mean(beta_1_hats)
beta_0_std <- sd(beta_0_hats)
beta_1_std <- sd(beta_1_hats)

beta_0_std_true <- sqrt(sigma ^ 2 / Sxx)
beta_1_std_true <- sqrt(sigma ^ 2 * (1 / 2000 + mean(x) ^ 2 / Sxx))

beta_0_col <- c(-5, beta_0_mean, beta_0_std_true, beta_0_std)
beta_1_col <- c(3.25, beta_1_mean, beta_1_std_true, beta_1_std)
summary_df <- data.frame(beta_0_col, beta_1_col)

```

**(c)** Plot two histograms side-by-side:

- A histogram of your simulated values for $\hat{\beta}_0$. Add the normal curve for the true sampling distribution of $\hat{\beta}_0$.
- A histogram of your simulated values for $\hat{\beta}_1$. Add the normal curve for the true sampling distribution of $\hat{\beta}_1$.

```{r}
hist(beta_0_hats, prob = TRUE, breaks = 20, 
     xlab = expression(hat(beta)[0]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_0_mean, sd = beta_0_std_true), 
      col = "darkorange", add = TRUE, lwd = 3)

hist(beta_1_hats, prob = TRUE, breaks = 20, 
     xlab = expression(hat(beta)[1]), main = "", border = "dodgerblue")
curve(dnorm(x, mean = beta_1_mean, sd = beta_1_std_true), 
      col = "darkorange", add = TRUE, lwd = 3)
```
***

## Exercise 4 (Simulating Confidence Intervals)

For this exercise we will simulate data from the following model:

\[
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\]

Where $\epsilon_i \sim N(0, \sigma^2).$ Also, the parameters are known to be:

- $\beta_0 = 5$
- $\beta_1 = 2$
- $\sigma^2 = 9$

We will use samples of size $n = 25$.

Our goal here is to use simulation to verify that the confidence intervals really do have their stated confidence level. Do **not** use the `confint()` function for this entire exercise.

**(a)** Simulate this model $2500$ times. Each time use `lm()` to fit a simple linear regression model, then store the value of $\hat{\beta}_1$ and $s_e$. Set a seed using **your** birthday before performing the simulation. Note, we are simulating the $x$ values once, and then they remain fixed for the remainder of the exercise.

```{r}
birthday = 19880210
set.seed(birthday)
n = 25
x = seq(0, 2.5, length = n)
```

```{r}
beta_1_hat_seq = rep(0, times=2500)
s_e_seq = rep(0, times = 2500)

for(i in 1:2500) {
  sim_res <- sim_slr(x = x, beta_0 = 5, beta_1 = 2, sigma = 3)
  linear_model <- lm(response ~ predictor, data = sim_res)
  

  beta_1_hat_seq[i] <- summary(linear_model)$coefficients["predictor", "Estimate"]
  s_e_seq[i] <- summary(linear_model)$coefficients["predictor", "Std. Error"]
}

```


**(b)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 95% confidence interval. Store the lower limits in a vector `lower_95` and the upper limits in a vector `upper_95`. Some hints:

- You will need to use `qt()` to calculate the critical value, which will be the same for each interval.
- Remember that `x` is fixed, so $S_{xx}$ will be the same for each interval.
- You could, but do not need to write a `for` loop. Remember vectorized operations.

```{r}
crit = qt(0.95, df = 25 - 2)
lower_95 <- beta_1_hat_seq - crit * s_e_seq
upper_95 <- beta_1_hat_seq + crit * s_e_seq
```


**(c)** What proportion of these intervals contains the true value of $\beta_1$?
```{r}
beta_1 <- 2
interval_df <- data.frame(lower_95, upper_95)
interval_df <- interval_df[interval_df$lower_95 <= 2,]
interval_df <- interval_df[interval_df$upper_95 >= 2,]
nrow(interval_df)
```


**(d)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.05$?

```{r}
2243/2500
```

**(e)** For each of the $\hat{\beta}_1$ that you simulated, calculate a 99% confidence interval. Store the lower limits in a vector `lower_99` and the upper limits in a vector `upper_99`.
```{r}
crit = qt(0.99, df = 25 - 2)
lower_99 <- beta_1_hat_seq - crit * s_e_seq
upper_99 <- beta_1_hat_seq + crit * s_e_seq
interval_df <- data.frame(lower_99, upper_99)
interval_df <- interval_df[interval_df$lower_99 <= 2,]
interval_df <- interval_df[interval_df$upper_99 >= 2,]

```

**(f)** What proportion of these intervals contains the true value of $\beta_1$?

```{r}
nrow(interval_df)/2500
```
**(g)** Based on these intervals, what proportion of the simulations would reject the test $H_0: \beta_1 = 0$ vs $H_1: \beta_1 \neq 0$ at $\alpha = 0.01$?

0.9792

***

## Exercise 5 (Prediction Intervals "without" `predict`)

Write a function named `calc_pred_int` that performs calculates prediction intervals:

$$
\hat{y}(x) \pm t_{\alpha/2, n - 2} \cdot s_e\sqrt{1 + \frac{1}{n}+\frac{(x-\bar{x})^2}{S_{xx}}}.
$$

for the linear model

$$
Y_i = \beta_0 + \beta_1 x_i + \epsilon_i.
$$

**(a)** Write this function. You may use the `predict()` function, but you may **not** supply a value for the `level` argument of `predict()`. (You can certainly use `predict()` any way you would like in order to check your work.)

The function should take three inputs:

- `model`, a model object that is the result of fitting the SLR model with `lm()`
- `newdata`, a data frame with a single observation (row)
    - This data frame will need to have a variable (column) with the same name as the data used to fit `model`.
- `level`, the level (0.90, 0.95, etc) for the interval with a default value of `0.95`

The function should return a named vector with three elements:

- `estimate`, the midpoint of the interval
- `lower`, the lower bound of the interval
- `upper`, the upper bound of the interval

```{r}
calc_pred_int <- function (model, newdata, level = 0.95) {

  
  beta_hat_1 = coef(model)[[2]];
  beta_hat_0 = coef(model)[[1]];
  y_hat = beta_hat_0 + beta_hat_1 * x;
  e = resid(model);
  n = length(resid(model));
  x = as.vector(model$model[,2]);
  y = as.vector(model$model[,1]);
  x_bar = mean(x);
  
  estimate = beta_hat_0 + beta_hat_1 * newdata[[1]];
  se = summary(model)$coefficients[2, "Std. Error"]
  Sxx = sum((x - mean(x)) ^ 2)
  
  pred = se * sqrt(1 + (1/length(x)) + ((newdata[[1]] - x_bar)^2 / Sxx));
  c("estimate" = estimate, "lower" = estimate - pred, "upper" = estimate + pred);
}

```

**(b)** After writing the function, run this code:

```{r, eval = FALSE}
newcat_1 = data.frame(Bwt = 4.0)
calc_pred_int(cat_model, newcat_1)
```

**(c)** After writing the function, run this code:

```{r, eval = FALSE}
newcat_2 = data.frame(Bwt = 3.3)
calc_pred_int(cat_model, newcat_2, level = 0.90)
```


